Tracing the Meta-Level: PyPy's JIT Compiler

### 2009-04-01 talk/icooolps2009/paper.tex
fix/answer some of anto's comments

To improve this situation, the tracing JIT could trace the execution of several
opcodes, thus effectively unrolling the bytecode dispatch loop. Ideally, the
bytecode dispatch loop should be unrolled exactly so much, that the unrolled version
corresponds to \emph{user loop}. User loops
occur when the program counter of the language interpreter has the
same value many times. This program counter is typically stored in one or several
variables in the language interpreter, for example the bytecode object of the
currently executed function of the user program and the position of the current
bytecode within that.  In the example above, the program counter is represented by 
the \texttt{bytecode} and \texttt{pair} variables.
\anto{XXX: why ``many times''? Twice is enough to have a loop, though is not hot}

Since the tracing JIT cannot know which parts of the language interpreter are
the program counter, the author of the language interpreter needs to mark the
relevant variables of the language interpreter with the help of a \emph{hint}.
The tracing interpreter will then effectively add the values of these variables
to the position key. This means that the loop will only be considered to be
closed if these variables that are making up program counter at the language
interpreter level are the same a second time. \sout{Such a loop is a loop of the user
program.} \anto{Loops found in this way are, by definition, user loops}.

The program counter of the language interpreter can only be the same a
second time after an instruction in the user program sets it to an earlier
value. This happens only at backward jumps in the language interpreter. That
means that the tracing interpreter needs to check for a closed loop only when it
encounters a backward jump in the language interpreter. Again the tracing JIT
cannot known where the backward branch is located, so it needs to be told with
the help of a hint by the author of the language interpreter.
\anto{XXX: the backward jumps are in the user program, not in the language
  interprer. Am I missing something?}

@@

\sout{
There is a similar conceptual problem about the point where tracing is started.
Tracing starts when the tracing interpreter sees one particular loop often
enough. This loop is always going to be the bytecode dispatch loop of the
language interpreter, so the tracing interpreter will start tracing all the
time. This is not sensible. It makes more sense to start tracing only if a
particular loop in the user program would be seen often enough. Thus we
need to change the lightweight profiling to identify the loops of the user
program. Therefore profiling is also done at the backward branches of the
language interpreter, using one counter per seen program counter of the language
interpreter.
}

\anto{I find the previous para a bit confusing. What about something more
  lightweight, like the following?}

\anto{Similarly, the interpreter uses the same techniques to detect \emph{hot
    user loops}: the profiling is done at the backward branches of the user
  program, using one counter per seen program counter of the language
  interpreter.}

@@

\anto{XXX: they driver does not list \emph{all} the variables; e.g. \texttt{n}
  is not listed.  But maybe we can just ignore this issue}

@@

\anto{I propose to show also the trace with the malloc removal enabled, as it
  is much nicer to see. Maybe we can say that the experimental optimization we
  are working on would generate this and that}

@@

\anto{Once we get the highly optimized trace, we can pass it to the \emph{JIT
    backend}, which generates the correspondent machine code. XXX: do we want
  to say something more about backends?}

@@

% YYY (anto)

@@

\textbf{Trace Trees:} This paper ignored the problem of guards that fail in a
large percentage of cases because there are several equally likely paths through
a loop. Just falling back to interpretation in this case is not practicable.
\sout{
Therefore we also start tracing from guards that failed many times and produce
machine code for that path, instead of always falling back to interpretation. 
}
\anto{
Therefore, if we find a guard that fails often enough, we start tracing from
there and produce efficient machine code for that case, instead of alwayas
falling back to interpretation.
}

@@

\anto{XXX: should we say that virtualizables are very cool, that nobody else
  does that and that they are vital to get good performaces with python
  without sacrificing compatibility?}

>>>>

To improve this situation, the tracing JIT could trace the execution of several
opcodes, thus effectively unrolling the bytecode dispatch loop. Ideally, the
bytecode dispatch loop should be unrolled exactly so much, that the unrolled version
corresponds to \emph{user loop}. User loops
occur when the program counter of the language interpreter has the
same value several times. This program counter is typically stored in one or several
variables in the language interpreter, for example the bytecode object of the
currently executed function of the user program and the position of the current
bytecode within that.  In the example above, the program counter is represented by 
the \texttt{bytecode} and \texttt{pair} variables.

Since the tracing JIT cannot know which parts of the language interpreter are
the program counter, the author of the language interpreter needs to mark the
relevant variables of the language interpreter with the help of a \emph{hint}.
The tracing interpreter will then effectively add the values of these variables
to the position key. This means that the loop will only be considered to be
closed if these variables that are making up program counter at the language
interpreter level are the same a second time.  Loops found in this way are, by
definition, user loops.

The program counter of the language interpreter can only be the same a
second time after an instruction in the user program sets it to an earlier
value. This happens only at backward jumps in the language interpreter. That
means that the tracing interpreter needs to check for a closed loop only when it
encounters a backward jump in the language interpreter. Again the tracing JIT
cannot known which part of the language interpreter implements backward jumps,
so it needs to be told with the help of a hint by the author of the language
interpreter.

@@

The language interpreter uses a similar techniques to detect \emph{hot user
loops}: the profiling is done at the backward branches of the user program,
using one counter per seen program counter of the language interpreter.

@@

@@

\anto{XXX I propose to show also the trace with the malloc removal enabled, as it
  is much nicer to see. Maybe we can say that the experimental optimization we
  are working on would generate this and that} \cfbolz{This example is not about
  mallocs! There are no allocations in the loop. The fix would be to use
  maciek's lazy list stuff (or whatever it's called) which is disabled at the
  moment}

@@

Once we get this highly optimized trace, we can pass it to the \emph{JIT
backend}, which generates the correspondent machine code.

@@

@@

\textbf{Trace Trees:} This paper ignored the problem of guards that fail in a
large percentage of cases because there are several equally likely paths through
a loop. Just falling back to interpretation in this case is not practicable.
Therefore, if we find a guard that fails often enough, we start tracing from
there and produce efficient machine code for that case, instead of alwayas
falling back to interpretation.

@@

\anto{XXX: should we say that virtualizables are very cool, that nobody else
  does that and that they are vital to get good performaces with python
  without sacrificing compatibility?} \cfbolz{no: feels a bit dishonest to not
  describe them properly and then say that they are very cool and vital}
